Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-05 sections/tensors.tex
good throught tensors

The first three of these functions, $declare$, $freeze$, and $thaw$, have to do with managing when tensors can be assumed mutable or immutable.
%
As we use Looplets to represent iteration over a tensor, we must restrict the mutability of tensors while we iterate over them. 
%
For example, if a tensor declares it has a constant
region from $i = 2:5$, but some other part of the computation modifies the
tensor at $i = 3$, this would result in incorrect behavior.
%
It is much easier to
write correct Looplet code if we can assume that the tensor is immutable while
we are reading from it.
%
Thus, we introduce the notion that a tensor can be in
read-only mode or update-only mode.  
%
In read-only mode, the tensor may only
appear in the right-hand side of assignments
%
. In update-only mode, the tensor
may only appear in the left-hand side of an assignment, either being overwritten
or incremented by some operator. 
%
We can switch between these modes using freeze
and thaw functions.
%
The declare function is used to allocate a tensor,
initialize it to some specified size and value, and leave it in update-only
mode.

The unfurl function is used to manage iteration over a subfiber. 
%
At the point when it comes time to iterate over a tensor, be in on the left or right hand
side of an assignment, we call the unfurl function to precompute whatever state
and datastructures are necessary to return a looplet nest that would iterate
over that level of a tensor.
%
The unfurl function is called directly before
iterating over the corresponding loop, so it has access to any state variables introduced
by freezing or thawing the tensor.

@@

Our view of a level as a fiber allocator implies an allocation function
$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation
function, instead relying on initialization to reset the fiber if it needs to be
reused. While all of the previous functions are used to manage the lifecycle and
iteration over a general tensor, the assemble function is quite specific to the
level abstraction, and the notion of positions within sublevels. Note: it was an
intentional choice to hold the parent level responsible for managing the
data of the sublevels, which positions they allocate, etc. This allows the parent
level to reuse allocation logic from internal index datastructures. For example,
a sparse level might use a list of indices to store which nonzeros are present,
and when it comes time to resize that list, it could also call assemble to resize the
sublevel, reducing the number of branches in the code. The assemble function
lends itself particularly to a "vector doubling" allocation approach, which we
have found to be effective and flexible when managing the allocation
of sparse left hand sides. This benefit is made clear in our evaluation section,
where we see that prior systems like TACO do not support all possible loop
orderings and format combinations for sparse matrix multiply because they do
not have a flexible enough allocation strategy, instead using a two-phase approach
which requires computing a complicated closed-form kernel to iterate over the
data twice to determine the number of required output nonzeros.

@@

    In the Finch programming model, the programmer relies on the Finch compiler to specialize to the sequential properties of the data. In our experience, the main benefits of specializing to structure come from the following properties of the data:
    \begin{itemize}
        \item \textbf{Sparsity} Sparse data is data that is mostly zero, or some other
        fill value. When we specialize on this data, we can use annihilation ($x
        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
        redundant work.

@@

        \item \textbf{Singletons} When we have only one non-fill region in sparse data,
        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
    \end{itemize}
    In the following section, we consider a set of concrete implementations of levels that expose all combinations of these structures, paying some attention to a few important special cases: random access, scalars, and leaf levels. We summarize the structures in Table~\ref{table:formats} and Table~\ref{tab:TypesOfStructure}.

@@

\subsection{Implementations of Structures}
\subsubsection{Sequentially Constructed Levels}
   We consider all combinations of these structural properties in Table
    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
    the the 8 resulting situations. 
    While it is impossible to write code which
    precisely addresses every possible structure, the combination of our level
    formats can express a wide variety of hierarchical structures to a
    sufficient granularity that we can generate code which utilizes the four
    properties.  For example, though banded arrays are a superset of ragged
    arrays, representing a ragged array using a banded format does not add much
    overhead.  The structures we consider are exhaustive in the sense that they
    address all combinations of sparsity, blocks, runs, and singletons in each
    level. We can represent a wide variety of hierarchical tensor structures by
    combining these level structures in a tree, as shown in
    Figure~\ref{fig:structuraldiversity}.

\subsubsection{Non-sequentially Constructed Levels}
To reduce the implementation burden and improve efficiency in the common case, the levels we described in the previous section only support bulk,
sequential creation of formats. However, when users want to be able to write out
of order (which is a common requirement arising from loop order or from the
problem itself, it occurs in our SpGEMM algorithms and our histogram example in
the evaluation section), we must use more complicated datastructures like hash
tables and trees to support the random writes so that in-order levels can be constructed later. Because these datastructures are more complex and have a higher implementation burden and performance overhead,
we only support random access construction of sparse or dense structures.  We
can use these two more general structures as intermediates to convert to our
more specialized structures later.

\subsubsection{Scalars}
Because elements are geared towards representing many leaves in a level, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
%
Scalars don't have as much structure as clearly only two structures apply.
%
Through these structures, scalars can also affect other tensors in crucial ways.
%
Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
%
Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
%
We also provide early break scalars, which modify the stepper Looplets to
re-specialize the loop whenever a reduction into an early break scalar hits an
annihilator.
%
We assume that we don't need to re-specialize other Looplets because the stepper
is the only one which repeats a non-constant number of times.

\begin{wrapfigure}{r}{.5\textwidth}
        \centering
        \vspace{-20pt}
        \scriptsize
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \vspace{-8pt}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
        \vspace{-36pt}
    \end{wrapfigure}

\HIDE{\saman{Can we not put future work into the submission!} As future work, we also recommend the implementation of a tree-based format, as this would enable incremental construction. Currently, the cost of freezing and thawing is usually proportional to the number of stored fibers, so an operation that sets a single value in a tensor would be prohibitively expensive (i.e.  $A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an intermediate randomly accessible structure, because (unlike blocks or singletons), runs have asymptotic benefits. However, these cases were not necessary for our case studies.}

\help{this section fails to address freeze or thaw and how it relates to the concrete levels}

>>>>

The first three of these functions, $declare$, $freeze$, and $thaw$, have to do with managing when tensors can be assumed mutable or immutable.
%
As we use Looplets to represent iteration over a tensor, we must restrict the mutability of tensors while we iterate over them. 
%
For example, if a tensor declares it has a constant
region from $i = 2:5$, but some other part of the computation modifies the
tensor at $i = 3$, this would result in incorrect behavior.
%
It is much easier to
write correct Looplet code if we can assume that the tensor is immutable while
we are reading from it.
%
Thus, we introduce the notion that a tensor can be in
read-only mode or update-only mode.  
%
In read-only mode, the tensor may only
appear in the right-hand side of assignments
%
. In update-only mode, the tensor
may only appear in the left-hand side of an assignment, either being overwritten
or incremented by some operator. 
%
We can switch between these modes using freeze
and thaw functions.
%
The $declare$ function is used to allocate a tensor,
initialize it to some specified size and value, and leave it in update-only
mode.

The $unfurl$ function is used to manage iteration over a subfiber. 
%
At the point when it comes time to iterate over a tensor, be in on the left or right hand
side of an assignment, we call $unfurl$ to precompute whatever state
and datastructures are necessary to return a looplet nest that would iterate
over that level of a tensor.
%
We call $unfurl$ directly before
iterating over the corresponding loop, so it has access to any state variables introduced
by freezing or thawing the tensor.

@@

Our view of a level as a fiber allocator implies an allocation function
$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation
function, instead relying on initialization to reset the fiber if it needs to be
reused. While all of the previous functions are used to manage the lifecycle and
iteration over a general tensor, $assemble$ is quite specific to the
level abstraction, and the notion of positions within sublevels. Note: it was an
intentional choice to hold the parent level responsible for managing the
data of the sublevels, which positions they allocate, etc. This allows the parent
level to reuse allocation logic from internal index datastructures. For example,
a sparse level might use a list of indices to store which nonzeros are present,
and when it comes time to resize that list, it could also call $assemble$ to resize the
sublevel, reducing the number of branches in the code. The $assemble$ function
lends itself particularly to a "vector doubling" allocation approach, which we
have found to be effective and flexible when managing the allocation
of sparse left hand sides. This benefit is made clear in our case studies,
where prior systems like TACO do not support all possible loop
orderings and format combinations for sparse matrix multiply because they do
not have a flexible enough allocation strategy, instead using a two-phase approach
which requires computing a complicated closed-form kernel to iterate over the
data twice to determine the number of required output nonzeros.

@@

\begin{wrapfigure}{r}{.4\textwidth}
    \vspace{-12pt}
    \centering
    \scriptsize
    \begin{tabular}{|c|c|c|c|l|}
        \hline
        \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singular } & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\ \hline
        &  &  &  & Dense \\ \hline
        &  &  & $\checkmark$ & n/a \\ \hline
        &  & $\checkmark$ &  & DenseRLE \\ \hline
        &  & $\checkmark$ & $\checkmark$ & n/a \\ \hline
        & $\checkmark$ &  &  & n/a \\ \hline
        & $\checkmark$ &  & $\checkmark$ & n/a \\ \hline
        & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
        & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
       $\checkmark$ &  &  &  & Sparse \\ \hline
       $\checkmark$ &  &  & $\checkmark$ & SparsePinpoint \\ \hline
       $\checkmark$ &  & $\checkmark$ &  & SparseRLE \\ \hline
       $\checkmark$ &  & $\checkmark$ & $\checkmark$ & SparseInterval \\ \hline
       $\checkmark$ & $\checkmark$ &  &  & SparseVBL \\ \hline
       $\checkmark$ & $\checkmark$ &  & $\checkmark$ & SparseBand \\ \hline
       $\checkmark$ & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
       $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
    \end{tabular}
    \caption{All combinations of our 4 structural properties and the
    corresponding formats we have chosen to represent them. Not all combinations
    are relevant. Note that blocks and runs need not be considered together
    because we must store a run length for each run, so there isn't a
    significant storage benefit to combining them. Blocks and singletons only
    make sense in the contex of sparsity.}
    \vspace{-24pt}
    %We omit such combinations from the exhaustive table.
    \label{tab:TypesOfStructure}
\end{wrapfigure}

    In the Finch programming model, the programmer relies on the Finch compiler to specialize to the sequential properties of the data. In our experience, the main benefits of specializing to structure come from the following properties of the data:

    \begin{itemize}
        \item \textbf{Sparsity} Sparse data is data that is mostly zero, or some other
        fill value. When we specialize on this data, we can use annihilation ($x
        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
        redundant work.

@@

        \item \textbf{Singular} When we have only one non-fill region in sparse data,
        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
    \end{itemize}
    In the following section, we consider a set of concrete implementations of levels that expose all combinations of these structures, paying some attention to a few important special cases: random access, scalars, and leaf levels. We summarize the structures in Table~\ref{table:formats} and Table~\ref{tab:TypesOfStructure}.

@@

\subsection{Implementations of Structures}
\subsubsection{Sequentially Constructed Levels}
   We consider all combinations of these structural properties in Table
    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
    the the 8 resulting situations. 
    While it is impossible to write code which
    precisely addresses every possible structure, our level
    formats can be combined to express a wide variety of hierarchical structures to a
    sufficient granularity that we can generate code which utilizes the four
    properties.  For example, though banded arrays are a superset of ragged
    arrays, representing a ragged array using a banded format does not add much
    overhead.  The structures we consider are exhaustive in the sense that they
    address all combinations of sparsity, blocks, runs, and singletons in each
    level. We can represent a wide variety of hierarchical tensor structures by
    combining these level structures in a tree, as shown in
    Figure~\ref{fig:structuraldiversity}.

\subsubsection{Non-sequentially Constructed Levels}
To reduce the implementation burden and improve efficiency in the common case, the levels we described in the previous section only support bulk,
sequential construction of formats. However, when users want to be able to write out
of order (which is a common requirement arising from loop order or from the
problem itself, it occurs in our SpGEMM algorithms and our histogram example in
the evaluation section), we must use more complicated datastructures like hash
tables and trees to support the random writes so that in-order levels can be constructed later. Because these datastructures are more complex and have a higher implementation burden and performance overhead,
we only support random access construction of sparse or dense structures.  We
can use these two more general structures as intermediates to convert to our
more specialized structures later.

\subsubsection{Scalars}
Because leaf levels are geared towards representing multiple leaves, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
%
Scalars don't have as much structure as clearly only two structures apply.
%
Through these structures, scalars can also affect other tensors in crucial ways.
%
Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
%
Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
%
We also provide early break scalars, which modify the stepper Looplets to
re-specialize the loop whenever a reduction into an early break scalar hits an
annihilator.
%
We assume that we don't need to re-specialize other Looplets because the stepper
is the only one which repeats a non-constant number of times.

%\HIDE{\saman{Can we not put future work into the submission!} As future work, we also recommend the implementation of a tree-based format, as this would enable incremental construction. Currently, the cost of freezing and thawing is usually proportional to the number of stored fibers, so an operation that sets a single value in a tensor would be prohibitively expensive (i.e.  $A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an intermediate randomly accessible structure, because (unlike blocks or singletons), runs have asymptotic benefits. However, these cases were not necessary for our case studies.}

%\help{this section fails to address freeze or thaw and how it relates to the concrete levels}
